Avoiding tipping points in fisheries management through Gaussian Process Dynamic Programming

### 2014-11-07 manuscripts/manuscript.tex
update data file names to match figures

Decision making under uncertainty is a ubiquitous challenge in the
management of human intervention in natural resources and conservation.
Ecological dynamics are frequently complex and difficult to measure,
making uncertainty in our understanding a persistent challenge.
Decision-theoretic approaches provide a framework to determine the best
sequence of actions in face of uncertainty, but only when that
uncertainty can be meaningfully quantified (Fischer et al. 2009). Over
the last four decades (but stretching even further back in time), marked
by the seminal contributions of Clark (1976), Clark (2009) and Walters
and Hilborn (1978), dynamic optimization methods, particularly
Stochastic Dynamic Programming (SDP), have become increasingly important
as a means of understanding how to manage human intervention into
natural systems. Simultaneously, there has been increasing recognition
of the importance of Allee population sizes or `tipping points'
(Scheffer et al. 2001, Polasky et al. 2011) that management actions may
inadvertently cause populations to cross. Scheffer et al. (2009)
emphasized the difficulties of formulating even qualitatively correct
models of the underlying processes.

We develop a novel approach to address these concerns in the context of
fisheries; though the underlying challenges and methods are germane to
many other problems of conservation or natural resource exploitation,
such as forestry. Economic value and ecological concern have made marine
fisheries the crucible for much of the founding work (Gordon 1954, Clark
1976, 2009, May et al. 1979, Reed 1979, Ludwig and Walters 1982) on
management under uncertainty. Global declines (Worm et al. 2006) and the
controversy surrounding their interpretation (Hilborn 2007, Worm et al.
2009) make understanding these challenges all the more pressing.

Uncertainty enters the decision-making process at many levels: intrinsic
stochasticity in biological processes, measurements, and implementation
of policy (\emph{e.g.} Reed 1979, Clark and Kirkwood 1986, Roughgarden
and Smith 1996, Sethi et al. 2005), parametric uncertainty (\emph{e.g.}
Ludwig and Walters 1982, Hilborn and Mangel 1997, McAllister 1998,
Schapaugh and Tyre 2013), and model or structural uncertainty
(\emph{e.g.} Williams 2001, Cressie et al. 2009, Athanassoglou and
Xepapadeas 2012). Of these, structural uncertainty is generally the
hardest to quantify. Typical approaches assume a weak notion of model
uncertainty in which the correct model (or reasonable approximation) of
the dynamics must be identified from among a handful of alternative
models, using either model choice, model averaging, or introducing yet
greater model complexity of which others may be special cases (model
averaging being one such way to construct such a model) (Williams 2001,
Cressie et al. 2009, Athanassoglou and Xepapadeas 2012). Even setting
aside other computational and statistical concerns (e.g. Cressie et al.
(2009)), these approaches do not address our second concern -
representing uncertainty outside the observed data range.

Model uncertainty is particularly insidious when model predictions must
be made outside of the range of data on which the model was estimated,
and models may be abused (Mangel et al. 2001). Extrapolation uncertainty
is felt most keenly in decision-theoretic (or optimal control)
applications, since (a) exploring the space of all possible actions
involves considering actions that may move the system outside the range
of observed behavior, and (b) decision-theoretic algorithms rely not
only on reasonable estimates of the expected outcomes, but depend on the
weights given to all possible outcomes (\emph{e.g.} Weitzman 2013),
hence estimates of the uncertainty can be as important as estimates of
the average or expected outcome.

Concerns about tipping points in ecological dynamics (Scheffer et al.
2001, Polasky et al. 2011) highlight this problem because they arise
from this very situation. In such cases, precise models are not
available and data is limited or biased to a particular region of state
space, such as high stock levels or an otherwise desirable state for the
ecosystem. If such models or data were available, one would know just
how far a system could be pushed before going over the tipping point,
and management would be simple. The fact that this concern is frequently
discussed highlights how often we have only limited data and inperfect
models. Moreover, the case of tipping points highlights how even small
errors can have very large consequences, as we shall illustrate later.

The dual concerns of model uncertainty and incomplete data coverage pose
a substantial challenge to existing decision-theoretic approaches
(Brozović and Schlenker 2011). Because intervention is often (but not
always, see Hughes et al. (2013)) too late after a tipping point has
been crossed, management is most often concerned with avoiding
potentially catastrophic tipping points before any data are available at
or following a transition that would more clearly reveal these regime
shift dynamics (e.g. Bestelmeyer et al. 2012).

We illustrate how a Stochastic Dynamic Programming (SDP) algorithm
(Mangel and Clark 1988, Marescot et al. 2013) can be driven by the
predictions from a Bayesian Non-Parametric (BNP) model of population
dynamics (Munch et al. 2005a). This provides two distinct advantages
compared with contemporary approaches. First, using a BNP sidesteps the
need for an accurate model-based description of the system dynamics.
Second, the BNP can better reflect uncertainty that arises when
extrapolating a model outside of the data on which it was fit. We
illustrate that when the correct model is not known, this latter feature
is crucial to providing a robust decision-theoretic approach in face of
substantial structural uncertainty. We use \emph{robust} to distinguish
approaches that provide nearly optimal solutions (as measured by the
chosen value function) without being sensitive to the choice of the
(unknown) underlying model.

This paper represents the first ecological application of the SDP
decision-making framework without an a priori model of the underlying
dynamics. In contrast to parametric models which can only reflect
uncertainty in parameter estimates, the BNP approach provides a
state-space dependent representation of uncertainty. This permits a much
greater uncertainty far from the observed data than near the observed
data. These features allow the Gaussian Process Dynamic Programming
(GPDP) approach to find robust management solutions in face of limited
data and without knowledge of the correct model structure.

To illustrate GPDP, we consider the performance of the GPDP policy
against the policies derived under several alternative parametric models
(Reed 1979, Ludwig and Walters 1982, Mangel and Clark 1988). The nature
of decision-making problems provides a compelling and pragmatic way to
compare models: Rather than compare models in terms of best fit to data,
we define model performance in the concrete terms of the
decision-maker's objectives.

@@

We first describe the requirements of dynamic optimization for the
management of human intervention in natural resource systems. After that
we describe three parametric models for population dynamics and the
Gaussian Process (GP)\footnote{Note that we abbreviate Gaussian Process
  as GP, which refers only to the statistical model we use to
  approximate the population dynamics (as opposed to a familiar
  parametric model such as the Ricker equation), while we use the term
  GPDP, Gaussian Process Dynamic Programming, to refer to the \emph{use}
  of a GP as the underlying process model when solving a Dynamic
  Program, as we illustrate in this paper. Hence we will refer to the
  models as: GP, Ricker, Allen, etc, while we refer to the novel method
  we put forward here as GPDP.} description of population dynamics.

@@

Dynamic optimization requires characterizing the dynamics of a state
variable (or variables), a control action, and a value function. In this
paper, we consider only a single state variable, so that the ideas can
be described as simply as possible. That is, we focus on a class of
one-dimensional models of population dynamics as a `best-case' scenario
for the parametric model performance relative to the GP. This is a
best-case scenario for the parametric models because the underlying
dynamics are always simulated from one of the parametric models in the
set, whereas in reality we never have the ``true'' model among our
candidates. By choosing one-dimensional models with few parameters, we
limit the chance that poor performance will be due to our inability to
estimate parameters well, something that becomes a more severe problem
for higher-dimensional parametric models. Further, the parametric models
we consider are those most commonly used in modeling stock-recruitment
dynamics or to model sudden transitions between alternative stable
states.

@@

We let $x(t)$ denote the size (numbers or biomass) of the focal
population at time $t$ and assume that in the absence of take its
dynamics are. (Throughout, we will use $X(t+1)$ with a capital $X$ as a
reminder that this is a stochastic, uncertainty quantity at time $t$,
while we use lowercase $x(t)$ as a reminder that the current population
size is treated as a given without error. The latter assumption is
straight-forward to relax but will simplify the presentation here).

\begin{equation}
X(t+1) = Z(t) f(x(t), p) \label{eq1} 
\end{equation}

Where Z(t) is log-normally distributed process stochasticity (Reed 1979)
and $p$ is a vector of parameters to be estimated from the data. We
describe three choices for $f(X(t),p)$ in the next section.In this
simple model, the control action is a harvest or take, $h(t)$, measured
in the same units as $X$, at time $t$. Thus, in the presence of take,
the population size on the right hand side of Eqn 1 is replaced by
$S(t) = X(t) - h(t)$.

To construct the value function, we consider a return when $X(t) = x $
and harvest $h(t) = h$ denoted as the reward, $R(x, h)$. For example, if
the return is simply the harvest at time $t$, then
$R(x, h) = \min(x, h)$. We assume that future harvests are discounted
relative to current ones at a constant rate of discount $\delta$ and ask
for the harvest policy that maximizes total discounted harvest between
the current time $t$ and a final time $T$. That is, we seek to maximize
over choices of harvest
$\mathbf{E}_{X(t+1)} [ \sum_{t = 0}^{T}  R(X(t), h(t), t) \delta^t]$,
where the state dynamics are given by Eqn 1 and $\mathbf{E}$ denotes the
expectation over the process stochasticity in the future population
state.

In order to find that policy, we introduce the value function
$V(x(t), t)$ representing the total discounted catch from time $t$
onwards given that $X(t) = x(t)$. This value function satisfies an
equation of SDP (Mangel and Clark 1988, Clark and Mangel 2000, Clark
2009), sometimes called the Hamilton-Jacobi-Bellman equation (Mangel
2014),

@@

We consider three candidate parametric models of the stock-recruitment
dynamics: The Ricker model, the Allen model (Allen and Tanner 2005), the
Myers model (Myers et al. 1995), Eqns \eqref{ricker}-\eqref{myers},
which we parameterize as follows for $f(S(t),p)$ in Eqn 1. In all three,
we let $K$ denote the carrying capacity and $r$ the maximum per capita
growth rate. In the Ricker model,

\begin{equation}
f(S(t)|p) = S(t) e^{r \left(1 - \frac{S(t)}{K} \right) } \label{ricker}
\end{equation}

In the Allen model,

\begin{equation}
f(S(t)|p) = S(t) e^{r \left(1 - \frac{S(t)}{K}\right)\left(S(t) - X_c\right)} \label{allen}
\end{equation}

Where $r$ and $K$ are as before, and $X_c$ denotes the location of the
unstable steady state (i.e., the tipping point).

In the Myers model,

\begin{equation}
f(S(t) | p)  = \frac{r S(t)^{\theta}}{1 + \frac{S(t)^\theta}{K}} \label{myers}
\end{equation}

Where $\theta = 1$ corresponding to Beverton-Holt dynamics and $\theta$
\textgreater{} 2 leads to Allee effects and multiple stable states.

The Ricker model involves two parameters, corresponding to a growth rate
and a carrying capacity, and cannot support alternative stable growth
rate dynamics. The Allen model resembles the Ricker dynamics with an
added Allee effect parameter (Courchamp et al. 2008), below which the
population cannot persist. The Myers model also has three parameters and
contains an Allee threshold, but has compensatory rather than
over-compensatory density dependence (resembling a Beverton-Holt curve
rather than a Ricker curve at high densities.)

The multiplicative log-normal stochasticity perturbs the growth
predicted by each of these deterministic model skeletons. This
introduces one additional parameter $\sigma$ that must be estimated.

Because of our interest in management performance in the presence of
tipping points, all of our simulations are based on the Allen model. The
Allen model is thus ``structurally correct'' and is expected to provide
a best-case scenario when the `true' dynamics are known. The Ricker
model is a reasonable approximation of these dynamics far from the Allee
threshold (but lacks threshold dynamics), while the Myers model shares
the essential feature of a threshold but differs in the structure.

Throughout the presentation, we will refer to the use of the ``True''
model to denote solving for the optimal solution when the underlying
parameters \emph{are known without error}, and refer to the ``Allen''
model to denote the when these parameters have instead only been
estimated from the sample data.

We consider a period of 40 years of training data: long enough that the
estimates do not depend on the particular realization, while longer
times are not likely to provide substantial improvement.\\Each of the
models is fit to the same training data (Figure 1).

@@

The core difference for our purpose between the GP representation and
the parametric models is that the GP model is defined explicitly in
reference to some observed data. Unlike the models above, it cannot be
specified by the value of some parameters alone. Although called
nonparametric, the GP (a) still involves the estimation of parameters,
and (b), statisticians also use non-parametric to mean models that make
only weak distributional assumptions (e.g. Lehmann 1975). However, we
retain the terminology for consistency with previous studies.

The use of GP methods to formulate a predictive model is relatively new
in the context of modeling dynamical systems (Kocijan et al. 2005), and
was first introduced in the context ecological modeling and fisheries
management in Munch et al. (2005b). GP models have subsequently been
used to test for the presence of Allee effects (Sugeno and Munch 2013a),
estimate the maximum reproductive rate (Sugeno and Munch 2013b),
determine temporal variation in food availability (Sigourney et al.
2012), and provide a basis for identifying model-misspecification
(Thorson et al. 2014). An accessible and thorough introduction to the
formulation and use of GPs can be found in Rasmussen and Williams
(2006).

A GP is an infinite dimensional (because it is a function of a
continuous input variable) stochastic process. Any realization of the
process for some finite sample of points $n$ is a multivariate normal
distribution of dimension $n$. Thus, to characterize the GP we need a
mean function and a covariance function. We proceed as follows.

We assume that the data $X(t)$ are observed with some process
stochasticity,

@@

where $\varepsilon$ are IID normal random variables with zero-mean and
variance $\sigma_n^2$. Note that we have chosen to assume additive
stochasticity. While we could just as easily consider log-normal
stochasticity as in the parametric models, we make this choice to bring
home the point that the Gaussian process approach need not have
structurally correct noise form either.

We consider the fase where we have some previous observations of the
\emph{timeseries}, $X_{\textrm{obs}}(t)$, and wish to \emph{predict}
future values $X_{\textrm{p}}(t)$. We consider the observations as pairs
of points $X_{\textrm{obs}}(t), X_{\textrm{obs}}(t+1)$ for
$t \in [1,2, ..., T]$ up through observed time $T$. Under the GP, a
finite sample of these predictions at $n$ points into the future,
$X_{\textrm p}(\tau)$ for $\tau \in [T+1, T+2, ..., T+n]$, is simply
given by a multivariate normal ($\mathcal{N}$):

@@

which has mean and covariance matrix given by:

@@

Where $K$ is the chosen covariance kernel. (Compare to Eqn 2.22 of
Rasmussen and Williams (2006) for a more detailed derivation.) Many
options are possible for this kernel. Throughout, we use the radial
basis function for simplicity, a matrix whose $(i,j)$th element is given
by:

\begin{equation}
K_{i,j}(x_i, y_j) = \exp\left(\frac{-|\mathbf{x_i}-\mathbf{j}|^2}{2 \ell^2} \right)
\end{equation}

We use inverse Gamma priors on both the length-scale $\ell$ and
$\sigma$, of the form

\begin{equation}
f(x; \alpha, \beta) = \frac{\beta^\alpha}{\Gamma(\alpha)} x^{-\alpha - 1}\exp\left(-\frac{\beta}{x}\right)
\end{equation}

For the $\sigma$ prior, $\alpha = $ 5 and $\beta = $ 5. For $\ell$
prior, $\alpha = $ 10 and $\beta = $ 10.

Using the simulated training data we also estimate a Gaussian process
defined by a radial basis function kernel of two parameters: $\ell$,
which gives the characteristic length-scale over which correlation
between two points in state-space decays, and $\sigma$, which gives the
scale of the process noise by which observations $Y(t+1)$ may differ
from their predicted values $X(t+1)$ given an observation of the
previous state, $x(t)$ (see Munch et al. 2005a).

We use a Metropolis-Hastings Markov Chain Monte Carlo (Gelman et al.
(2003)) to infer posterior distributions of the two parameters of the GP
(Figure S3, code in appendix). Since the posterior distributions differ
substantially from the priors (Figure S3), we can be assured that most
of the information in the posterior comes from the data rather than the
prior belief.

@@

It is reasonably straight-forward to derive the harvest policy from the
estimated GP by inserting it into a SDP algorithm. The only difference
is that the uncertainty in the future state under the GP, $X(t+1)$,
includes both process uncertainty (based on the estimation of $\sigma$)
and ``structural uncertainty'' of the posterior collection of
curves.\\(Recall that given the parameters, a parametric model is
represented by a single curve, while the GP remains a distribution of
curves).\\From the GP posteriors, we can write down the (discritized)
transition matrix representing the probability of going to each state
$X(t+1)$ given any current state $x(t)$ and any harvest $h(t)$ (See the
function \texttt{gp\_transition\_matrix()} in the provided R package).
Given this transition matrix, we use the same value iteration algorithm
as in the parametric case to determine the optimal policy.

@@

All models fit the observed data rather closely and with relatively
small uncertainty, as illustrated in the posterior predictive curves in
Figure 1, in which we show the training data of stock sizes observed
over time as points, overlaid with the step-ahead predictions of each
estimated model using the parameters sampled from their posterior
distributions. Each model manages to fit the observed data rather
closely. Compared to the true model most estimates appear to over-fit,
predicting fluctuations that are actually due purely to stochasticity.
Model selection criteria (Table 1) penalize more complex models and show
a preference for the simpler Ricker model over the models with
alternative stable states (Allen and Myers). Details on MCMC estimates
for each model, traces, and posterior distributions can be found in the
appendix.

\begin{longtable}[c]{@{}cccc@{}}
\toprule\addlinespace
\begin{minipage}[b]{0.12\columnwidth}\centering
~
\end{minipage} & \begin{minipage}[b]{0.10\columnwidth}\centering
Allen
\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\centering
Ricker
\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\centering
Myers
\end{minipage}
\\\addlinespace
\midrule\endhead
\begin{minipage}[t]{0.12\columnwidth}\centering
\textbf{DIC}
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\centering
50.75
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
50.45
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
50.41
\end{minipage}
\\\addlinespace
\begin{minipage}[t]{0.12\columnwidth}\centering
\textbf{AIC}
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\centering
-24.51
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-30.13
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-27.01
\end{minipage}
\\\addlinespace
\begin{minipage}[t]{0.12\columnwidth}\centering
\textbf{BIC}
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\centering
-17.75
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-25.06
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-20.25
\end{minipage}
\\\addlinespace
\bottomrule
\addlinespace
\caption{Model selection scores for several common criteria (DIC:
Deviance Information Criterion, AIC: Akaike Information Criterion, BIC:
Bayesian Information Criterion) all (wrongly) select the simplest model.
As the true (Allen) model is not distinguishable from the simpler
(Ricker) model in the region of the observed data, this error cannot be
avoided regardless of the model choice criterion. This highlights the
danger of model choice when the selected model will be used outside of
the observed range of the data.}
\end{longtable}

@@

Hilborn, R. 2007. Reinterpreting the State of Fisheries and their
Management. Ecosystems 10:1362--1369.

@@

Lehmann, E. L. 1975. Nonparametrics: Statistical Methods Based on Ranks.
Holden-Day, IncSan Francisco.

>>>>

Decision making under uncertainty is a ubiquitous challenge in the
management of human intervention in natural resources and conservation.
Decision-theoretic approaches provide a framework to determine the best
sequence of actions in face of uncertainty, but only when that
uncertainty can be meaningfully quantified (Fischer et al. 2009). Over
the last four decades (beginning with Clark (1976), Clark (2009) and
Walters and Hilborn (1978)) dynamic optimization methods, particularly
Stochastic Dynamic Programming (SDP), have become increasingly important
as a means of understanding how to manage human intervention into
natural systems. Simultaneously, there has been increasing recognition
of the importance of multiple steady states or `tipping points'
(Scheffer et al. 2001, 2009, Polasky et al. 2011) in ecological systems.

We develop a novel approach to address these concerns in the context of
fisheries; although the challenges and methods are germane to other
problems of conservation or natural resource exploitation. Economic
value and ecological concern have made marine fisheries the crucible for
much of the founding work on management under uncertainty (Gordon 1954,
Clark 1976, 2009, May et al. 1979, Reed 1979, Ludwig and Walters 1982)

Even if we know the proper deterministic description of the biological
system, there is intrinsic stochasticity in biological dynamics,
measurements, and implementation of policy (\emph{e.g.} Reed 1979, Clark
and Kirkwood 1986, Roughgarden and Smith 1996, Sethi et al. 2005). We
may also lack knowledge about the parameters of the biological dynamics
(parametric uncertainty, \emph{e.g.} Ludwig and Walters 1982, Hilborn
and Mangel 1997, McAllister 1998, Schapaugh and Tyre 2013), or even not
know which model is proper description of the system (structural
uncertainty, \emph{e.g.} Williams 2001, Cressie et al. 2009,
Athanassoglou and Xepapadeas 2012). Of these, the latter is generally
the hardest to quantify. Typical approaches confront the data with a
collection of models, assuming that the true dynamics (or reasonable
approximation) is among the collection and then use model choice or
model averaging to arrive at a conclusion (Williams 2001, Cressie et al.
2009, Athanassoglou and Xepapadeas 2012). Even setting aside other
concerns (see Cressie et al. (2009)), these approaches are unable to
describe uncertainty outside the observed data range.

Structural uncertainty is particularly insidious when we try to predict
outside of the range of observed data (Mangel et al. 2001) because we
are extrapolating into unknown regions. In management applications, this
extrapolation uncertainty is particularly important since (a) management
involves considering actions that may move the system outside the range
of observed behavior, and (b) the decision tools (optimal control
theory, SDP) rely on both reasonable estimates of the expected outcomes
and on the weights given to those outcomes (\emph{e.g.} Weitzman 2013).
Thus characterizing uncertainty is as important as characterizing the
expected outcome.

Tipping points in ecological dynamics (Scheffer et al. 2001, Polasky et
al. 2011) highlight this problem because precise models are not
available and data are limited such as around high stock levels or an
otherwise desirable state. With perfect information , one would know
just how far a system could be pushed before crossing the tipping point,
and management would be simple. But we face imperfect models and limited
data and, with tipping points,\\even small errors can have very large
consequences, as we shall illustrate later. Because intervention is
often (but not always, see Hughes et al. (2013)) too late after a
tipping point has been crossed, management is often concerned with
avoiding tipping points before any data about them is available or the
transition that would more clearly reveal a tipping point occurs (e.g.
Bestelmeyer et al. 2012).

The dual concerns of model uncertainty and incomplete data create a
substantial challenge to existing decision-theoretic approaches
(Brozović and Schlenker 2011). We illustrate how Stochastic Dynamic
Programming (SDP) (Mangel and Clark 1988, Marescot et al. 2013) can be
implemented using a Bayesian Non-Parametric (BNP) model of population
dynamics (Munch et al. 2005a). The BNP method has two distinct
advantages . First, using a BNP model sidesteps the need for an accurate
model-based description of the system dynamics. Second, a BNP model
better reflects uncertainty when extrapolating beyond the observed data.
This is crucial to providing robust decision-making when the correct
model is not known (as is almost always true). {[}We use \emph{robust}
to characterize approaches that provide nearly optimal solutions without
being sensitive to the choice of the (unknown) underlying model.{]}

This paper is the first ecological application of the SDP without an
\emph{a priori} model of the underlying dynamics. Unlike parametric
approaches that can only reflect uncertainty in parameter estimates, the
BNP method provides a broader representation of uncertainty, including
uncertainty beyond the observed data. We will show that Gaussian Process
Dynamic Programming (GPDP) allows us to find robust management solutions
in face of limited data without knowing the correct model structure.

For comparisons, we consider the performance of management based on GPDP
against management policies derived under several alternative parametric
models (Reed 1979, Ludwig and Walters 1982, Mangel and Clark 1988).
Rather than compare models in terms of best fit to data, we compare
model performance in the concrete terms of the decision-maker's
objectives.

@@

We first describe the requirements of dynamic optimization for the
management of human intervention in natural resource systems. After that
we describe three parametric models for population dynamics and the
Gaussian Process (GP)\footnote{We abbreviate Gaussian Process as GP,
  which refers to the statistical model we use to approximate the
  population dynamics , and we use the term Gaussian Process Dynamic
  Programming {[}GPDP,{]}, to refer to the \emph{use} of a GP as the
  underlying process model when solving a Dynamic Programming equation .
  Hence we will refer to the models as: GP, Ricker, Allen, etc, and the
  novel method we put forward here as GPDP.} description of population
dynamics.

@@

Dynamic optimization requires characterizing the dynamics of a state
variable (or variables), a control action, and a value function. For
simplicity, we consider only a single state variable, which allows us to
focus on one-dimensional models of population dynamics for the
comparison of parametric models and the GP. This is a best-case scenario
for the parametric models because we simulate underlying dynamics from
one of the three parametric models , whereas in the natural world we
never know the ``true'' mode . In addition, by choosing one-dimensional
models with just a few parameters, we limit the chance that poor
performance will be due to inability to estimate parameters accurately,
something that becomes a more severe problem for higher-dimensional
parametric models. Finally, the parametric models we consider are those
most commonly used in modeling stock-recruitment dynamics or to model
sudden transitions between alternative stable states.

@@

We let $X(t)$ denote the size (numbers or biomass) of the focal
population at time $t$ and assume that in the absence of take its
dynamics are. .

\begin{equation}
X(t+1) = Z(t) f(X(t), p) \label{eq1} 
\end{equation}

Where Z(t) is log-normally distributed process stochasticity (Reed 1979)
and $p$ is a vector of parameters to be estimated from the data. We
describe the three choices for $f(X(t),p)$ in the next section.\\The
control action is a harvest or take, $h(t)$, measured in the same units
as $X$, at time $t$. Thus, in the presence of take, the population size
on the right hand side of Eqn 1 is replaced by $S(t) = X(t) - h(t)$.

To construct the value function, we consider a return when $X(t) = x(t)$
and harvest $h(t) = h$ denoted as the reward, $R(x(t), h)$. For example,
if the return is the harvest at time $t$, then
$R(x(t), h(t)) = \min(x(t), h(t))$. We assume that future harvests are
discounted relative to current ones at a constant rate of discount
$\delta$ and ask for the harvest policy that maximizes total discounted
harvest between the current time $t$ and a final time $T$. That is, we
seek to maximize over choices of harvest
$E_{X(t+1)} [ \sum_{t = 0}^{T}  R(X(t), h(t), t) \delta^t]$, where the
state dynamics are given by Eqn 1 and $E$ denotes the expectation over
the process stochasticity in the future population state.

In order to find that policy, we introduce the value function
$V(x(t), t)$ representing the total discounted catch from time $t$
onwards given that $X(t) = x(t)$. This value function satisfies an
equation of SDP (Mangel and Clark 1988, Clark and Mangel 2000, Clark
2009 Mangel 2014),

@@

We consider three candidate parametric models for the population
dynamics: The Ricker model, the Allen model (Allen and Tanner 2005), and
the Myers model (Myers et al. 1995), Eqns \eqref{ricker}-\eqref{myers}.
In all three, we let $K$ denote the carrying capacity and $r$ the
maximum per capita growth rate. The Ricker model has two parameters and
the right hand side of Eqn 1 is

\begin{equation}
f(S(t)|r,K) = S(t) e^{r \left(1 - \frac{S(t)}{K} \right) } \label{ricker}
\end{equation}

The Allen model has three parameters

\begin{equation}
f(S(t)|r, K, X_c) = S(t) e^{r \left(1 - \frac{S(t)}{K}\right)\left(S(t) - X_c\right)} \label{allen}
\end{equation}

where $X_c$ denotes the location of the unstable steady state (i.e., the
tipping point).

The Myers model also has three parameters

\begin{equation}
f(S(t) | r, K, \theta)  = \frac{r S(t)^{\theta}}{1 + \frac{S(t)^\theta}{K}} \label{myers}
\end{equation}

where $\theta = 1$ corresponds to Beverton-Holt dynamics and $\theta$
\textgreater{} 2 leads to Allee effects and multiple stable states.

The Ricker model does not lead to multiple steady states. The Allen
model resembles the Ricker dynamics with an added Allee effect parameter
(Courchamp et al. 2008), below which the population cannot persist. The
Myers model also has three parameters and contains an Allee threshold,
but unlike the Ricker model saturates at high population size. The
multiplicative log-normal stochasticity in Eqn 1 introduces one
additional parameter $\sigma$ that must be estimated.

Because of our interest in management performance in the presence of
tipping points, all of our simulations are based on the Allen model. The
Allen model is thus the state of nature and is expected to provide the
best-case scenario. The Ricker model is a reasonable approximation of
these dynamics far from the Allee threshold (but lacks threshold
dynamics), while the Myers model shares the essential feature of a
threshold but differs in structure from the Allen model. Throughout, we
refer to the ``True'' model when the underlying parameters \emph{are
known without error}, and refer to the ``Allen'' model when these
parameters have been estimated from the sample data.

We consider a period of 40 in which data are obtained to estimate the
parameters or the GP. This is long enough that the estimates do not
depend on the particular realization, and longer times are not likely to
provide substantial improvement.\\We fit each of the models is fit to
the same data (Figure 1).

@@

The core difference for our purpose between the GP and the parametric
models is that the GP model is defined explicitly by the observed data.
That is, it cannot be specified by the value of parameters alone.

The use of GPs to characterize dynamical systems is relatively new
(Kocijan et al. 2005), and was first introduced in the context
ecological modeling and fisheries management in Munch et al. (2005b). GP
models have subsequently been used to test for the presence of Allee
effects (Sugeno and Munch 2013a), estimate the maximum reproductive rate
(Sugeno and Munch 2013b), determine temporal variation in food
availability (Sigourney et al. 2012), and provide a basis for
identifying model-misspecification (Thorson et al. 2014). An accessible
and thorough introduction to the formulation and use of GPs can be found
in Rasmussen and Williams (2006).

A GP is a stochastic process for which any realization of for a finite
sample of points $n$ is a multivariate normal distribution of dimension
$n$. Thus, to characterize the GP we need a mean function and a
covariance function. We proceed as follows.

As before, we assume that the data $X(t)$ are observed with process
stochasticity around a mean function \$ g(X(t))\$

@@

where $\varepsilon$ are IID normal random variables with zero-mean and
variance $\sigma_n^2$. Note that we have chosen to assume additive
stochasticity. While we could consider log-normal stochasticity as in
the parametric models, we make this choice to emphasize that the
Gaussian process approach need not have structurally correct
stochasticity to be effective.

Now consider the situation where we have a set of previous observations
of the \emph{timeseries}, $X_{\textrm{obs}}(t)$, and wish to
\emph{predict} future values $X_{\textrm{p}}(t)$. We consider the
observations as pairs of points
$X_{\textrm{obs}}(t), X_{\textrm{obs}}(t+1)$ for $t=1, 2, …$ up through
final observation time $T$. As described above, in the GP, a finite
sample of these predictions at $n$ points into the future,
$X_{\textrm p}(\tau)$ for $\tau \in [T+1, T+2, ..., T+n]$, is a
multivariate normal ($\mathcal{N}$):

@@

with mean and covariance matrix given by (see Eqn 2.22 of Rasmussen and
Williams (2006) for a more detailed derivation)

@@

Where $K$, called the covariance kernel, measures how much influence one
observation has on another. In this paper, for $K$ use a matrix whose
$(i,j)$th element is given by:

\begin{equation}
K_{i,j}(x_i, y_j) = \exp\left( \frac{ -(x_i - y_j)^2}{2 \ell^2} \right)
\end{equation}

so that $\ell$ gives the characteristic length-scale over which
correlation between two observations decays.

We use inverse Gamma priors on both the length-scale $\ell$ and
$\sigma$, thus for example

\begin{equation}
f(\ell; \alpha, \beta) = \frac{\beta^\alpha}{\Gamma(\alpha)} \ell^{-\alpha - 1}\exp\left(-\frac{\beta}{\ell}\right)
\end{equation}

For the prior on $\ell$, $\alpha = $ 10 and $\beta = $ 10. The prior on
$\sigma$, $\alpha = $ 5 and $\beta = $ 5.

We use a Metropolis-Hastings Markov Chain Monte Carlo (Gelman et al.
(2003)) to infer posterior distributions of the parameters of the GP
(Figure S3, code in appendix). Since the posterior distributions differ
substantially from the priors (Figure S3), most of the information in
the posterior comes from the data rather than the prior belief.

@@

We derive the harvest policy from the estimated GP by inserting it into
a SDP algorithm. Given the GP posteriors, we construct the transition
matrix representing the probability of going to each state $X(t+1)$
given any current state $x(t)$ and any harvest $h(t)$ (See the function
\texttt{gp\_transition\_matrix()} in the provided R package). Given this
transition matrix, we use the same value iteration algorithm as in the
parametric case to determine the optimal policy.\\In doing so, the
uncertainty in the future state under the GP, $X(t+1)$, includes both
process uncertainty (based on the estimation of $\sigma$) and structural
uncertainty of the posterior collection of curves.\\Unlike the case of a
parametric model, which is a single curve given the parameters, the GP
remains a distribution of curves.

@@

All of the models fit the observed data rather closely and with
relatively small uncertainty. In Figure 1, we show the posterior
predictive curves. The training data of stock sizes observed over time
are points, overlaid with the step-ahead predictions of each estimated
model using the parameters sampled from their posterior distributions.
Compared to the true model most estimates appear to over-fit, predicting
patterns that are actually due purely to stochasticity. Model selection
criteria (Table 1) penalize more complex models and show a preference
for the simpler Ricker model over the models with alternative stable
states (Allen and Myers). Details on MCMC estimates for each model,
traces, and posterior distributions can be found in the appendix.

\begin{longtable}[c]{@{}cccc@{}}
\toprule\addlinespace
\begin{minipage}[b]{0.12\columnwidth}\centering
~
\end{minipage} & \begin{minipage}[b]{0.10\columnwidth}\centering
Allen
\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\centering
Ricker
\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\centering
Myers
\end{minipage}
\\\addlinespace
\midrule\endhead
\begin{minipage}[t]{0.12\columnwidth}\centering
\textbf{DIC}
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\centering
50.75
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
50.45
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
50.41
\end{minipage}
\\\addlinespace
\begin{minipage}[t]{0.12\columnwidth}\centering
\textbf{AIC}
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\centering
-24.51
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-30.13
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-27.01
\end{minipage}
\\\addlinespace
\begin{minipage}[t]{0.12\columnwidth}\centering
\textbf{BIC}
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\centering
-17.75
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-25.06
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-20.25
\end{minipage}
\\\addlinespace
\bottomrule
\addlinespace
\caption{Model selection scores for several common criteria (DIC:
Deviance Information Criterion, AIC: Akaike Information Criterion, BIC:
Bayesian Information Criterion) all select the wrong model. As the true
(Allen) model is not distinguishable from the simpler (Ricker) model in
the region of the observed data, this error cannot be avoided regardless
of the model choice criterion. This highlights the danger of model
choice when the selected model will be used outside of the observed
range of the data.}
\end{longtable}

@@

@@
